---
title: "metabolomics analysis"
author: "Duygu Dede Sener, Amaury Pelzer, Denise Slenter"
date: "2023-06-05"
output: html_document
---
## download and install required packages
```{r}
#install.packages('metabolighteR')
#or we can install it via the following code
remotes::install_github('aberHRML/metabolighteR')
library(metabolighteR)
library(tidyverse)
library(dplyr)
library(httr)
library(stringr)
if(!"ggplot2" %in% installed.packages()){install.packages("ggplot2")}
library('ggplot2')
library(ggrepel)
library(rJava)
library(jsonlite)
library(clusterProfiler)
#mtbls_key()
```

## Retrieve the data from Metaboligths data source
```{r}
BASE_URL <- "https://www.ebi.ac.uk:443/metabolights/ws"
study <- 'MTBLS3233' # this study ID can be replaced by the dataset ID captured in Fairspace

#get all available files for the study
allFiles <- metabolighteR::get_study_files(study, raw_data = FALSE)
#to get file/files ends with "maf.tsv"
selectedFiles <- allFiles[grepl("maf.tsv", allFiles$file),]

#the first and second contains same samples but different peak intensity values 
#for now we will only do analysis for the first profiling file
query <- paste0(BASE_URL, "/studies/",study,"/download?file=", selectedFiles$file[1])
metabolites <-  httr::GET (url = query, httr::add_headers(user_token = getOption('MTBLS_API_KEY'))) %>% 
                    content(as = "text") %>% 
                    read_delim(delim = "\t") %>% 
                    as_tibble()

# following is for the second profiling data, uncomment it when needed
# query2 <- paste0(BASE_URL, "/studies/",study,"/download?file=", selectedFiles$file[2])
# metabolites2 <-  httr::GET (url = query2, httr::add_headers(user_token = getOption('MTBLS_API_KEY'))) %>%
#                    content(as = "text") %>%
#                    read_delim(delim = "\t") %>%
#                    as_tibble()

#to get metadata
metaFile <- allFiles [allFiles$type=="metadata_sample",]
query <- paste0(BASE_URL, "/studies/",study,"/download?file=", metaFile$file)
metadata <-  httr::GET (url = query, httr::add_headers(user_token = getOption('MTBLS_API_KEY'))) %>% 
                    content(as = "text") %>% 
                    read_delim(delim = "\t") %>% 
                    as_tibble()

```

## Modify the retrieved data (metadata and count data)
```{r}
#drop NA values for factor value
#metadata <- metadata %>% drop_na (`Factor Value[Disease]`, )

#remove NA values from the column containing "Factor" string, since the column "Factor" would be the compared column
metadata <- metadata %>% filter(rowSums(is.na(.[ , grep("Factor", colnames(.))])) != ncol(.[ , grep("Factor", colnames(.))]))

#take only two columns
metadata <- metadata[,grepl("Factor", colnames(metadata)) | grepl("Sample Name", colnames(metadata))]
#metadata <- metadata[ ,c("Factor Value[Disease]", "Sample Name")] # remove disease, check sample name 
colnames(metadata) <- c("Sample","Factor") 

#filter out columns for metabolites data
columns <- metabolites %>% dplyr::select(database_identifier,chemical_formula,metabolite_identification)

#take only samples in metabolites occurring in the metadata
names.use <- names(metabolites)[ names(metabolites) %in% metadata$Sample]
#update mbx data with used sample names
metabolites <- metabolites [ ,names.use]

#order data based on column names
metabolites <- metabolites[ , order(names(metabolites),decreasing=TRUE)]
#add first selected description columns again
metabolites <- cbind(columns, metabolites)
#remove rows containing "unknown" values for database identifier column
metabolites <- metabolites[!grepl("unknown", metabolites$database_identifier),]
#remove rows containing NA values for database identifier column
metabolites <- metabolites %>% drop_na (`database_identifier`, )

#order metadata based on sample name
metadata <- metadata[ order(metadata$Sample,decreasing=TRUE),]

rm(allFiles,columns,metaFile, selectedFiles, BASE_URL,names.use, query, study)
```

## Preprocessing metabolomics data before statistical analysis
```{r}
#Remove metabolites has NA value for more than 50%
columns <- ncol(metabolites)
rows <- nrow(metabolites)
#detect how many NA values exist per metabolite
nofNAperMet <- rowSums(is.na(metabolites[,4:columns]))
#calculate fifty percent of total columns
fifty_percent <- floor((columns)/2)
mSet_MissingDataCounted <- cbind(metabolites, nofNAperMet)
mSet_NoMissingData <- subset(mSet_MissingDataCounted, nofNAperMet <= fifty_percent)
#Remove last column for further processing.
mSet_NoMissingData <- subset(mSet_NoMissingData, select=-c(nofNAperMet))

#Convert intensity data to numeric values                         
mSet_NoMissingData[, c(4:columns)] <- apply(mSet_NoMissingData[, c(4:columns)],2, function(x) as.numeric(as.character(x)))
transformation <- "log_2" 

if(transformation == "cube_root"){
    mSet_transformed <-  mSet_NoMissingData[, c(4:columns)]^(1/3)
}else if(transformation == "square_root"){
    mSet_transformed <-  mSet_NoMissingData[, c(4:columns)]^(1/2)
}else if(transformation == "log_2"){
    mSet_transformed <-  log2(mSet_NoMissingData[, c(4:columns)])
}else if(transformation == "log_10"){
    mSet_transformed <- log10(mSet_NoMissingData[, c(4:columns)])
}else{print("Warning: name for transformation not recognized")}

#add description columns to the transformed data
mSet_transformed <- cbind(mSet_NoMissingData[,c(1:3)], mSet_transformed)

write.csv(mSet_transformed$database_identifier,"IDs.csv", row.names = FALSE, quote = FALSE)
```

## Visualize and test whether the transformation is good for the transformed data
```{r}
## Visualize the data after the transformation (for one sample to get an idea of suitability of transformation:
#create histogram for original distribution for first column with data
hist(mSet_NoMissingData[,4], col='steelblue', main='Original')
#create histogram for log-transformed distribution 
hist(mSet_transformed[,4], col='coral2', main=transformation)

## Testing if the transformation creates a normally distributed dataset (alpha >= 0.05)
##Calculate all Shapiro values for raw and transformed data:
mSet_NoMissingData_Shapiro <- lapply(mSet_NoMissingData[,c(4:columns)], shapiro.test)
mSet_transformed_Shapiro <- lapply(mSet_transformed[,c(4:columns)], shapiro.test)

#Obtain the p-values for raw and transformed data
mSet_NoMissingData_Shapiro_pvalues <- do.call(rbind, mSet_NoMissingData_Shapiro)
mSet_transformed_Shapiro_pvalues <- do.call(rbind, mSet_transformed_Shapiro)

## Count how often the p-value is above 0.05, to obtain an estimate of achieved normality due to transformation
mSet_NoMissingData_Shapiro_pvalues_sum <- sum(mSet_NoMissingData_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)
mSet_transformed_Shapiro_pvalues_sum <- sum(mSet_transformed_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)
eighty_percent <- floor(((columns)/10)*8)

#Print relevant information:
if(mSet_transformed_Shapiro_pvalues_sum[1] > eighty_percent ){paste0("Data after ", transformation ," transformation seems to follow a normal distribution for more then 80% of your data")} else{
  print("Advised to select a different data transformation procedure")}

remove( mSet_MissingDataCounted, mSet_NoMissingData, mSet_NoMissingData_Shapiro, mSet_NoMissingData_Shapiro_pvalues, mSet_transformed_Shapiro, mSet_transformed_Shapiro_pvalues, eighty_percent, fifty_percent, mSet_NoMissingData_Shapiro_pvalues_sum, mSet_transformed_Shapiro_pvalues_sum)

```

## Statistical analysis
```{r}

#Create backup of data
mSet_transformed.b <- mSet_transformed
mSet_FINAL <- mSet_transformed

#get the first disease 
factor <- colnames(mSet_FINAL)[c(4:columns)]
#Find relevant columns per group.
columns_factor <- sum(str_count(colnames(mSet_FINAL)[c(4:columns)], factor))
end_factor <- columns_factor

##calculate logFC for 2 groups, ignoring missing values (NAs) when calculating the mean.  
group1 = apply(mSet_FINAL[,4:end_factor], 1, mean, na.rm=TRUE)
group2 = apply(mSet_FINAL[,(end_factor+1):ncol(mSet_FINAL)], 1, mean, na.rm=TRUE)


          # ###########  DESEQ implementation ###############
          # sampleLabels<-sampleLabels[, c(2,3)]
          # sampleLabels$disease <- relevel(factor(sampleLabels$disease),ref="nonIBD")
          # #add an experimental group variable to sampleLabels
          # sampleLabels$group <- as.factor(paste(sampleLabels$disease,sampleLabels$biopsy_location,sep="_"))
          # 
          # library(DESeq2)
          # sampleLabels <- metadata$Sample
          # dds <- DESeq2::DESeqDataSetFromMatrix(countData = data, colData=sampleLabels)
          # 
          # library(limma)
          # results <- limma::voom(counts = mSet_NoMissingData[,4:ncol(mSet_NoMissingData)], design = NULL, lib.size = NULL, normalize.method = "none",
          #      block = NULL, correlation = NULL, weights = NULL,
          #      span = 0.5, plot = FALSE, save.plot = FALSE)
          # 
          # v <- voom(counts= mSet_NoMissingData[,4:ncol(mSet_NoMissingData)], design, plot=TRUE, normalize="quantile")
          # 
          # ##################################################

#once the metabolomics data is log2 transformed, we need to take the difference between the means 
#(iso dividing the means over one another), since log2 Fold Change or log2 Ratio == log2(condition / control). 
#Note: if the transformation step applied is cube_root or square_root, one needs to divide control over disease for this step!
  if(transformation == "log2"){
    foldchange <-  group1 - group2
  } else if(transformation == "log10"){
    foldchange <-  log2((10^group1)/(10^group2))
  } else {
    foldchange<- log2(group1 /group2 )
  }

mSet_AnalysisReady <- cbind(mSet_FINAL, foldchange)

##Calculate p-value for two groups based on t-test
##general function to store p-values for multiple rows:
ttest_mSet <- function(df, grp1, grp2) {
  x = df[grp1]
  y = df[grp2]
  x = as.numeric(x)
  y = as.numeric(y)  
  results = t.test(x, y)
  results$p.value
}
p_values <- apply(mSet_FINAL, 1, ttest_mSet, grp1 = c(4:end_factor), grp2 = c((end_factor+1):ncol(mSet_FINAL)))

##Add p_values column to analysis dataset:
mSet_AnalysisReady <- cbind(mSet_AnalysisReady, p_values)

#Convert logFC and p-values columns to numeric values            
mSet_AnalysisReady <- as.data.frame(mSet_AnalysisReady)

mSet_AnalysisReady[ , c('foldchange','p_values')] <- apply(mSet_AnalysisReady[ , c('foldchange','p_values')], 2, function(x) as.numeric(as.character(x)))
```

## Volcano Plot to visualize the statistical analysis results 
```{r}

##Inspired by: https://biocorecrg.github.io/CRG_RIntroduction/volcano-plots.html
##Define the thresholds for log2 (Fold Change) and p-values
#For cut-off value uncertainties, see https://doi.org/10.1039/C6AN01342B .
log2FC_min <- -0.26
log2FC_max <-  0.26
p_value_threshold <- 0.05

##Create column with HMDB_IDs, only if the data is relevant
mSet_AnalysisReady$relevant_labels <- mSet_AnalysisReady$database_identifier
mSet_AnalysisReady$relevant_labels[!((mSet_AnalysisReady$foldchange <= log2FC_min 
                                    | mSet_AnalysisReady$foldchange >= log2FC_max) 
                                    &  mSet_AnalysisReady$p_values <= p_value_threshold)] <- NA

# ##Duplication issues:
# if(!"dplyr" %in% installed.packages()){install.packages("dplyr")}
# library(dplyr)
# library(ggrepel)

##volcanoPlot_Disorder 
volcanoPlot_disorder <- ggplot(data=mSet_AnalysisReady, aes(x=foldchange, y=-log10(p_values), 
                                                            label=relevant_labels)) + geom_point() + 
                                                            theme_minimal() + geom_text_repel()

## Add vertical lines for FoldChange and P-value thresholds:
volcanoPlot_disorder <- volcanoPlot_disorder + geom_vline(xintercept=c(log2FC_min, log2FC_max), col="blue") +
    geom_hline(yintercept=-log10(p_value_threshold), col="red") + theme(plot.background = element_rect(fill = "white"))

disorderName <- "Factor"

titleVolcano <- paste0("Volcano plot of ", transformation, " transformed data for ", disorderName )
verticalAxisTitle <- paste0(transformation, " Fold Change, ", disorderName, " versus control ")

## Add title and update axis labels:
volcanoPlot_disorder <- volcanoPlot_disorder + ggtitle(titleVolcano) + labs(y = "-log10(p-value)", x = verticalAxisTitle)

# Show the Volcano plot in the notebook output:
volcanoPlot_disorder

```


## Select statistically significant enzymes
```{r}
# Select statistically significant enzymes
sigMets <- mSet_AnalysisReady[which(p_values<0.05),]
sigMets$pValue <- p_values[which(p_values<0.05)]

```


# Get pathway data with compound transformation
The following Shinyapp was used to map HMDB IDs to KEGG IDs
https://tabbassidaloii.shinyapps.io/BridgeDb-Shiny/

## BridgeDB identifier mapping
## it does not work properly need to check again failed to create mapper object!
```{r}
##Set the working directory to download the Metabolite mapping file
location <- "data/metabolites_20220707.bridge"
checkfile <- paste0(getwd(), '/' ,location)

##Download the Metabolite mapping file (if it doesn't exist locally yet):
if (!file.exists(checkfile)) {
download.file(
  "https://figshare.com/ndownloader/files/26001794",
  location)
}
##Load the metabolite mapping file:
mapper <- BridgeDbR ::loadDatabase(checkfile)
#Load the ID mapper:
mapper <-  BridgeDbR ::loadDatabase(checkfile)
```


#To get mapped file for KEGG compound and CHEBI ids
```{r}
# Read KEGG compound to HMDB list
compound.CHEBI <- read.table(file = "data/XrefBatch_mapping_BridgeDB-Shiny.csv", sep = ',', header = TRUE)
# Remove ... columns
compound.CHEBI <- compound.CHEBI[, -c(2,4)]
# Change column names
colnames(compound.CHEBI) <- c("database_identifier", "compound")

#remove(mbxCount.log.CD, mbxCount.log.UC, mbxCount.log.nonIBD)
```


## Pathway Analysis
```{r}

# Merge the KEGG compound to the full metabolomics data
sigMets <- merge(sigMets, compound.CHEBI, by = "database_identifier", all.x = TRUE)
#length(which(is.na(deg.CD$compound)))
# remove NAs for compounds
sigMets <- sigMets[!is.na(sigMets$compound), ]
# Relocate the compound column as first
sigMets <- sigMets %>% relocate(compound)


```


## Calculate the ORA score for each pathway


### ORA score for Ulcerative colitis (UC)

```{r}

# Read JSON file with all the pathways and the compounds within into R object
PW.to.compound <- fromJSON("data/all_pathways_compounds_dict.json")
# Read JSON file with all the pathways codes and their names within into R object
PW.all.names <- fromJSON("data/ec_pathway_names_dict.json")

# Transform PW to compound for further analysis
PW.to.compound <- as.data.frame(t(as.data.frame(sapply(PW.to.compound, rbind))))
PW.to.compound$pathway <- rownames(PW.to.compound)
colnames(PW.to.compound) <- c("compound", "pathway")
PW.to.compound <- PW.to.compound[,c(2,1)]
PW.to.compound$pathway <- sub("\\..*", "", PW.to.compound$pathway)
PW.to.compound$compound <- sub("\\s\\s.*", "", PW.to.compound$compound)

PW.to.name <- data.frame("pathway"=names(PW.all.names), "name"=unlist(PW.all.names))

# Perform Pathway enrichment
res<- clusterProfiler::enricher(gene=as.factor(unlist(sigMets)),
                TERM2GENE = PW.to.compound,
                TERM2NAME = PW.to.name)

as.data.frame(res)
```

## Plot the results in a horizontal bar chart

```{r}
# Horizontal bar plot of significant pathways in UC
res[which(res$p.adjust < 0.05),] %>% 
  ggplot(aes(reorder(as.character(Description), Count), Count)) + 
  geom_col(aes(fill = p.adjust)) + 
  scale_fill_gradient2(low = "#0000ff", 
                       mid = "#ff0000",
                       high= "yellow",
                       midpoint = 0.001) +
  coord_flip() + 
  labs(x = "Pathways", y = "Significanlty changed metabolites count")
```
## Writing output tables in csv file

```{r}
# Write the output in a file
write.table(res, "output/mbxPWdata.csv", sep =",", row.names = FALSE)

```

## Creating jupyter notebook files
```{r}
#Jupyter Notebook file
if(!"devtools" %in% installed.packages()) BiocManager::install("devtools")
devtools::install_github("mkearney/rmd2jupyter", force=TRUE)
library(devtools)
library(rmd2jupyter)
rmd2jupyter("metabolomics_analysis.Rmd")
```


